---
title: 'Dates 1: Recognizing and Formatting Dates'
output:
  html_document:
    number_sections: true
    toc: true
    toc_float: true
    css: !expr here::here("global/style/style.css")
    highlight: kate
editor_options: 
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

```{r, echo = F, message = F, warning = F}
# Load packages 
if(!require(pacman)) install.packages("pacman")
pacman::p_load(rlang, tidyverse, knitr, here, reactable, gt, flextable)

## functions
source(here::here("global/functions/misc_functions.R"))

## default render
registerS3method("reactable_5_rows", "data.frame", reactable_5_rows)
knitr::opts_chunk$set(class.source = "tgc-code-block")
```

# Intro

Understanding how to manipulate dates is a crucial skill in health data analysis. From patient admission dates to vaccination schedules, date-related data plays a vital role in epidemiological analyses. In this lesson, we will learn how R stores and displays dates, as well as how to effectively manipulate, parse, and format them. Let's get started!

# Learning Objectives

‣You understand how dates are stored and manipulated in R

‣You understand how to coerce strings to dates

‣You know how to handle messy dates

‣You are able to change how dates are displayed

# Packages

Please load the packages needed for this lesson with the code below:

```{r warning = F, message = F, echo = T}
if(!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse, 
               lubridate)
```

## Datasets

‣ Today we'll work with **two datasets**.

‣ The first is about **Indoor Residual Spraying (IRS)** for **malaria control in Illovo, Malawi**.

```{r message=FALSE}
irs <- read_csv(here("data/Illovo_data.csv"))
irs
```

‣ The second is simulated data on **hospital inpatient stays**. It includes **admission and discharge dates** for 150 patients in different formats.

```{r message=FALSE}
ip <- read_csv(here("data/inpatient_data.csv"))
ip
```

------------------------------------------------------------------------

## Introduction to dates in R

‣ In R, we use the `Date` class to handle dates.

‣ Default format: **"YYYY-MM-DD"**

‣ If you just enter in such a date string, R will initially consider this to be a character:

```{r}
class("_____________")
```

‣ To specify a date, we can use `as.Date()` function.

```{r}
my_date <- as.Date("______________")
class(my_date)
```

------------------------------------------------------------------------

‣ With `Date`, we can find the **difference between two dates**.\

```{r}
as.Date("2020-01-01") - as.Date("2019-01-01")
```

This would of course not be possible if you had bare characters:

```{r eval = F}
"2020-01-01" - "2000-12-20"
```

------------------------------------------------------------------------

‣ The default format for `as.Date()` is "YYYY-MM-DD".

‣ Other common formats like "MM/DD/YYYY" or "Month DD, YYYY" won't work by default (later we'll see how to "parse" such dates):

```{r, eval=FALSE}
as.Date("12/31/2000") 
as.Date("Dec 31, 2000")
```

‣ However, R will also accept "/" instead of "-" as long as the order is still "YYYY/MM/DD".

```{r}
as.Date("2000/12/31"
```

# Coercing (parsing) strings to dates

‣ Let's focus on how R classifies date variables in our **IRS dataset**.

‣ We'll examine the columns containing dates and see how R reads them.

```{r}
irs %>% 
  select(contains("date")
```

------------------------------------------------------------------------

‣ To parse the remaining dates, we need to specify the date format. Why? Ambiguity.

‣ For example: `01-02-03`. Is it January 2nd, 2003, or February 1st, 2003?

------------------------------------------------------------------------

## Using Base R for Date coercion (parsing)

‣ As we saw, the function `as.Date()` is used to convert strings to dates in base R.\

‣ Let's try it without specifying the format of the date.

```{r}
irs %>% 
  select(start_date_typical) %>% 
  mutate(start_date_typical = as.Date(start_date_typical))
```

‣ As you see, not what we wanted. Original format was "DD/MM/YYYY".

------------------------------------------------------------------------

‣ To correctly interpret dates, we use the `format` argument in `as.Date()`

‣ Symbols to specify date components:

| Component                                          | Symbol | Example |
|----------------------------------------------------|--------|---------|
| Year (numeric, with century)                       | %Y     | 2023    |
| Year (numeric, without century)                    | %y     | 23      |
| Month (numeric, 01-12)                             | %m     | 01      |
| Month (written out fully)                          | %B     | January |
| Month (abbreviated)                                | %b     | Jan     |
| Day of the month                                   | %d     | 31      |
| Day of the week (numeric, 1-7 with Sunday being 1) | %u     | 5       |
| Day of the week (written out fully)                | %A     | Friday  |
| Day of the week (abbreviated)                      | %a     | Fri     |

‣ Let's try with `format = "%d%m%Y"`

```{r, render = reactable_5_rows}
irs %>%
  select(start_date_typical) %>% 
  mutate(start_date_typical = as.Date(start_date_typical, format = "%d%m%Y"))

```

‣ Still did not work? Why?

‣ Format needs to include the slashes `%d/%m/%Y`

```{r}
irs %>%
  select(start_date_typical) %>% 
  mutate(start_date_typical = as.Date(start_date_typical, format = "%d/%m/%Y"))
```

------------------------------------------------------------------------

**PRACTICE TIME !**

::: practice
**Coerce long date**

Try to coerce the column `start_date_long` from the IRS dataset to the `Date` class. Don't forget to include all elements into the format option, including the symbols that separate the components of the date!

```{r, eval = FALSE}
# Complete the code with your answer:
irs %>% 
  _____________________
```
:::

::: r-practice
**Find code errors**

Can you find all the errors in the following code?

```{r}
as.Date("June 26, 1987", format = "%b%d%y")
```
:::

# Lubridate

‣ The `lubridate` package offers a **user-friendly way** to convert strings to dates.

‣ Specify the order of **year**, **month**, and **day** using "y", "m", and "d".

‣ Separators between components are **not necessary**.

------------------------------------------------------------------------

```{r}
# Example conversions
dmy("8 Sep 2021")

mdy("Jan 01 2020")

class(ymd("2023 May 23")
```

------------------------------------------------------------------------

‣ Let's apply these functions to the `start_day_long` variable in our dataset.

```{r, render = reactable_5_rows}
irs %>%
  select(start_date_long) %>%  
  mutate(start_date_long_conv = mdy(start_date_long)) 
  
```

------------------------------------------------------------------------

**PRACTICE TIME !**

::: r-practice
**Coerce typical date**

Try to coerce the column `start_date_typical` from the IRS dataset to the `Date` class using the functions in the lubridate package.

```{r, eval = FALSE}
# Your code here
```
:::

------------------------------------------------------------------------

**PRACTICE TIME !**

::: r-practice
**Base and lubridate formatting**

The following table contains the formats found in the `adm_date_abbr` and `adm_date_messy` formats from our inpatient dataset. See if you can fill in the blank cells:

| Date example | Base R   | Lubridate |
|--------------|----------|-----------|
| Dec 07, 2022 |          |           |
| 03-27-2022   |          | mdy       |
| 28.04.2022   |          |           |
|              | %Y/%m/%d |           |
:::

------------------------------------------------------------------------

‣ We've learned methods to convert strings to dates.

‣ But what about dealing with **multiple date formats** in one column?

------------------------------------------------------------------------

## Handling Messy Dates with `lubridate::parse_date_time()`

‣ Working with dates often involves multiple formats within a single column.

‣ `lubridate` offers `parse_date_time()` function to handle such scenarios.

‣ The function uses the `orders` argument for specifying multiple date formats.

------------------------------------------------------------------------

‣ Let's start with a basic example:

```{r}
parse_date_time("30/07/2001", orders = "dmy")
```

‣ The function works well for single string inputs and is equivalent to using `dmy()` in this case.

------------------------------------------------------------------------

‣ The power of `parse_date_time()` is truly evident when handling multiple formats.

‣ Example with two different formats:

```{r}
parse_date_time(c("1 Jan 2000", "July 30th 2001"), orders = c("dmy", "mdy"))
```

‣ Order of formats in `orders` argument does not matter when the date strings are distinct:

```{r}
parse_date_time(c("1 Jan 2000", "July 30th 2001"), orders = c("mdy", "dmy"))
```

------------------------------------------------------------------------

‣ Ambiguity in dates can be a challenge.

‣ For ambiguous dates, the order in `orders` does matter.

```{r}
parse_date_time(c("01/02/2000", "01/03/2000"), orders=c("mdy", "dmy"))
```

```{r}
parse_date_time(c("01/02/2000", "01/03/2000"), orders=c("dmy", "mdy"))
```

------------------------------------------------------------------------

**PRACTICE TIME !**

::: r-practice
**Using parse_date_time**

The dates in the code below are November 9th 2002, December 4th 2001, and June 5th 2003. Complete the code to coerce them from strings to dates.

```{r echo=TRUE, eval=FALSE}
parse_date_time(c("11/09/2002", "12/04/2001", "2003-06-05"), orders=c(...))
```
:::

------------------------------------------------------------------------

‣ Returning to our dataset, focus on the `start_date_messy` column.

‣ We are dealing with four different formats:

\- YYYY/MM/DD

\- Month DD YYYY

\- DD-MM-YYYY

\- MM/DD/YYYY

‣ Now that we know how our data is formatted, we can use the `parse_date_time()` function to clean it up.

```{r render = kable_5_rows}
irs %>%
  select(start_date_messy) %>%
  mutate(start_date_clean = parse_date_time(start_date_messy, 
                                            orders = c("mdy", "dmy", "ymd")))
```

‣ Is order important here? Let's try a different order and find out!

```{r render = kable_5_rows}
irs %>%
  select(start_date_messy) %>%
  mutate(start_date_messy_1 = parse_date_time(start_date_messy, 
                                              orders = c("mdy", "dmy", "ymd")),
         start_date_messy_2 = parse_date_time(start_date_messy, 
                                              orders = c("dmy", "mdy", "ymd")))
```

‣ If you scroll through, you will be able to observe that the conversions were the same.

‣ Why? Because there is no ambiguity in the dataset, and the order of formats provided to `orders` only matter when we have ambiguous dates.

‣ You may think that there IS ambiguity: perhaps, the first date could be July 4, or April 7:

```{r}
irs %>%
  select(start_date_messy) %>% 
  slice(1)
```

‣ But {lubridate} read it as April 7 ("mdy") in both cases, even when "dmy" was provided first for the `start_date_messy_2` column?

‣ This is because there are other rows that **resolve** this amibiguity. For example, row 11 is "06/26/14":

```{r}
irs %>%
  select(start_date_messy) %>% 
  slice(1, 11)
```

‣ This row can ONLY be interpreted as June 26. The number 26 cannot represent a month.

‣ So lubridate knows that all rows of the format "XX/XX/XX" in the dataset are "mm/dd/yy"

::: r-practice
**Using parse_date_time with adm_date_messy**

With the help of the table you completed from the exercise, use the `parse_date_time()` function to clean up the `adm_date_messy` column in the inpatient dataset, `ip`!
:::

------------------------------------------------------------------------

## Changing how dates are displayed

‣ Up to now, we've been using the `Date` class with a **default "YYYY-MM-DD" format**.

‣ But, how can we **display dates in a specific format** for reports or graphs?

‣ We can use the `format()` function in base R.

‣ This function gives us the freedom to **customize the appearance of dates**.

------------------------------------------------------------------------

‣ Here's a table to remind us how different date parts are represented:

| Component                                          | Symbol | Example |
|----------------------------------------------------|--------|---------|
| Year (numeric, with century)                       | %Y     | 2023    |
| Year (numeric, without century)                    | %y     | 23      |
| Month (numeric, 01-12)                             | %m     | 01      |
| Month (written out fully)                          | %B     | January |
| Month (abbreviated)                                | %b     | Jan     |
| Day of the month                                   | %d     | 31      |
| Day of the week (numeric, 1-7 with Sunday being 1) | %u     | 5       |
| Day of the week (written out fully)                | %A     | Friday  |
| Day of the week (abbreviated)                      | %a     | Fri     |

------------------------------------------------------------------------

‣ Let's try a single example:

‣ We want to convert `2000-01-31` to "Jan 31, 2000".

```{r}
my_date <- as.Date("2000-01-31")
my_date
format(my_date, "%b %d, %Y")
```

------------------------------------------------------------------------

**PRACTICE TIME !**

::: practice
**Format date vector**

Format the date below to MM/DD/YYYY using the format function:

```{r}
my_date <- as.Date("2018-05-07")
```
:::

------------------------------------------------------------------------

‣ Now, let's apply this on our dataset:

‣ Creating a new variable `start_date_char` from `start_date_default`.

‣ We'll display it as **DD/MM/YYYY**.

```{r, render = reactable_5_rows}
irs %>% 
  select(start_date_default) %>% 
  mutate(start_date_char = format(start_date_default, "%d %b %Y")) %>% 
  select(start_date_default, start_date_char)
```

------------------------------------------------------------------------

‣ One last example:

‣ Using `end_date_default` and formatting it as **Month DD, YYYY**.

```{r, render = reactable_5_rows}
irs %>% 
  select(end_date_default) %>% 
  mutate(end_date_char = format(end_date_default, "%B %d, %Y")) %>% 
  select(end_date_default, end_date_char)
```

------------------------------------------------------------------------

## Wrap Up!

‣ Congratulations on finishing the first Dates lesson!

‣ You now understand **how Dates are stored, displayed, and formatted in R**.

‣ Next, we'll learn how to perform **manipulations with dates** and create **basic time series graphs**.

------------------------------------------------------------------------

# Answer Key {.unnumbered}

**Coerce long date**

```{r echo=TRUE, eval=FALSE}
irs <- irs %>%
  mutate(start_date_long = as.Date(start_date_long, format="%B %d %Y"))
```

**Find code errors**

```{r echo=TRUE, eval=FALSE}
as.Date("June 26, 1987", format = "%B %d, %Y")

```

**Coerce typical date**

```{r echo=TRUE, eval=FALSE}
irs %>%
  mutate(start_date_typical = dmy(start_date_typical))
```

**Base and lubridate formatting**

| Date example | Base R    | Lubridate |
|--------------|-----------|-----------|
| Dec 07, 2022 | %b %d, %Y | mdy       |
| 03-27-2022   | %m-%d-%Y  | mdy       |
| 28.04.2022   | %d.%m.%Y  | dmy       |
| 2021/05/23   | %Y/%m/%d  | ymd       |

**Using parse_date_time**

```{r echo=TRUE, eval=FALSE}
parse_date_time(c("11/09/2002", "12/04/2001", "2003-06-05"), orders=c("mdy", "ymd"))
```

**Using parse_date_time with adm_date_messy**

```{r echo=TRUE, eval=FALSE}
ip %>%
  mutate(adm_date_messy = parse_date_time(adm_date_messy, orders = c("mdy", "dmy", "ymd")))

```

**Format date vector**

```{r echo=TRUE, eval=FALSE}
my_date <- as.Date("2018-05-07")
format(my_date, "%m/%d/%Y")
```

------------------------------------------------------------------------

# Contributors {.unlisted .unnumbered}

The following team members contributed to this lesson:

`r tgc_contributors_list(c("amckinley", "kendavidn"))`