---
title: 'Data Cleaning Pipeline 2:  Fixing Inconsistencies'
output:
  html_document:
    number_sections: true
    toc: true
    toc_float: true
    css: !expr here::here("global/style/style.css")
    highlight: kate
editor_options: 
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

```{r, echo = F, message = F, warning = F}
if(!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse, knitr, here)

## functions
source(here::here("global/functions/misc_functions.R"))

## default render
registerS3method("reactable_5_rows", "data.frame", reactable_5_rows)
knitr::opts_chunk$set(class.source = "tgc-code-block")
```

# Introduction

In the previous lesson, we learned a range of functions for diagnosing data issues. Now, let's focus on some common techniques and functions for fixing those issues. Let's get started!

# Learning Objectives

By the end of this lesson, you will be able to:

-   Understand how to clean column names, both automatically and manually.
-   Effectively eliminate duplicate entries.
-   Correct and fix string values in your data.
-   Convert data types as required.

# Packages

Load the following packages for this lesson:

```{r warning = F, message = F, echo = F}
# Load packages 
if(!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse,
               janitor,
               inspectdf)
```

## Dataset

‣ Working with a **modified version** of the dataset from the first `Data Cleaning` lesson.

‣ **More errors** have been added for cleaning purposes.

```{r}
non_adherence <- read_csv(here("data/non_adherence_messy.csv"))
non_adherence
```

## Cleaning column names

‣ Column names should be **clean** and **standardized** for ease of use and readability.

‣ Ideal column names should be **short**, have **no spaces or periods**, **no unusual characters**, and **similar style**.

‣ Use the `names()` function from base R to check column names of our `non_adherence` dataset.

```{r}
 # check column names
```

‣ Some names have **spaces**, **special characters**, or are **not uniformly cased**.

## Automatic column name cleaning with `janitor::clean_names()`

‣ Use `janitor::clean_names()` to **standardize column names**.

```{r}
non_adherence %>%
  clean_names() %>%
  names()
```

‣ Observe changes like **upper case to lower case**, **spaces to underscores**, and **periods replaced**.

‣ Let's save this cleaned dataset as `non_adherence_clean`.

```{r}
________________ <- 
  ________________ %>%
  clean_names()
```

::: r-practice
### Q: Automatic cleaning {.unlisted .unnumbered}

*(NOTE: Answers are at the bottom of the page. Try to answer the questions yourself before checking.)*

The following dataset has been adapted from a study that used retrospective data to characterize the tmporal and spatial dynamics of typhoid fever epidemics in Kasene, Uganda.

```{r eval = F}
typhoid <- read_csv(here("data/typhoid_uganda.csv"))
```

Use the `clean_names()` function from `janitor` to clean the variables names in the `typhoid` dataset.
:::

## {stringr} and `dplyr::rename_with()` for Renaming Columns

‣ `rename_with()` from `dplyr` allows applying functions to all column names. Sometimes easier to use than `rename()`.

‣ Example: Convert all column names to upper case with `rename_with(colname, toupper)`.

```{r}
________________ %>%
  rename_with(________________)
```

‣ Another task: In the `non_adherence` dataset, remove `_of_patient` from column names for simplicity.

‣ Use `stringr::str_replace_all()` within `rename_with()` for this task.

‣ `str_replace_all()` syntax: `str_replace_all(string, pattern, replacement)`.

```{r}
test_string <- "this is a test test string" # replace test with new
str_replace_all(________________)
```

‣ Apply `str_replace_all()` to remove `_of_patient` in column names of `non_adherence_clean`.

```{r}
non_adherence_clean_2 <-  
  ________________ # non_adherence_clean then rename_with()
```

::: side-note
Remember, creating many intermediate objects like `non_adherence_clean` and `non_adherence_clean_2` is for tutorial clarity. In practice, combine multiple cleaning steps in a single pipe chain:

```{r eval = F}
non_adherence_clean <- 
  non_adherence %>%
  # cleaning step 1 %>%
  # cleaning step 2 %>%
  # cleaning step 3 %>%
  # etc.
```
:::

::: r-practice
### Q: Complete cleaning of column names {.unlisted .unnumbered}

Standardize the column names in the `typhoid` dataset with `clean_names()` then;

-   replace `or_` with `_`

-   replace `of` with `_`
:::

## Removing Duplicate Rows

‣ Duplicated rows in datasets can be due to **multiple data sources** or **survey responses**.

‣ It's **essential** to **identify and remove these duplicates** for accurate analysis.

‣ Use `janitor::get_dupes()` to **identify duplicate rows**. This allows for **visual inspection** before removal.

```{r}
 # Use get_dupes() to identify duplicates
```

‣ After identifying, use `dplyr::distinct()` to **remove duplicates**, keeping only the **unique rows**.

```{r}
# Before removal
nrow(________________)

# Removing duplicates
non_adherence_distinct <- 
  non_adherence_clean_2 %>% 
  ________________

# After removal
nrow(________________)
```

‣ Re-check for duplicates with `get_dupes()` to ensure all have been removed.

```{r}
________________ %>% 
  get_dupes()
```

::: r-practice
### Q: Removing duplicates {.unlisted .unnumbered}

Identify the duplicates in the `typhoid` dataset using `get_dupes()`, then remove them using `distinct()`.
:::

## Homogenize strings

‣ We observed **inconsistent capitalization** in string characters, like `Professor` and `professor`, in the `occupation` variable.

‣ To address this, we can **transform character columns to a specific case**. Here, we'll use **title case**. Preferable for graphics and reports.

```{r}
non_adherence_case_corrected <- 
  non_adherence_distinct # then the across function

  # check the values of age_35 and occupation

  # check the updated values of age_35 and occupation
```

::: r-practice
### Q: Transforming to lowercase {.unlisted .unnumbered}

Transform all the strings in the `typhoid` dataset to lowercase.
:::

## `dplyr::case_match()` for String Cleaning

‣ We will explore the `case_match()` function from the {dplyr} package for string cleaning.

‣ `case_match()` allows for specifying conditions and values to be applied to a vector.

‣ Here is an example using `case_match()`:

```{r eval = F}
test_vector <- c("+", "-", "NA", "missing")
case_match(_____) # + to positive, - to negative, default as unknown
```

‣ The function takes a vector and series of conditions. `.default` is optional for unmatched conditions.

‣ Let's apply `case_match()` to the `sex` column in the `non_adherence_distinct` dataset.

‣ First, observe the levels in this variable:

```{r}
non_adherence_distinct %>% count(sex)
```

‣ Inconsistencies in the `sex` column coding can be fixed using `case_match()`. Let's change `F` to `Female`:

```{r}
# case match F to Female, with default as is

```

‣ This function is useful for multiple value changes, like in the `occupation` column.

‣ Modifications to be made: - "Worker" to "Laborer" - "Housewife" to "Homemaker" - "Truck Driver" and "Taxi Driver" to "Driver"

```{r}
non_adherence_recoded <- 
  non_adherence_case_corrected %>%
  mutate(sex = case_match(sex, "F" ~ "Female", .default = sex)) %>% 
  # case match Worker to Laborer, Housewife to Homemaker, Truck Driver and Taxi Driver to Driver
  
```

::: warning
Remember to use `.default=column_name` in `case_match()`. Without it, unmatched values become `NA`.
:::

::: r-practice
### Q: Fixing strings {.unlisted .unnumbered}

The variable `householdmembers` from the `typhoid` dataset should represent the number of individuals in a household. There is a value `01-May` in this variable. Recode this value to `1-5`.
:::

## Converting Data Types

‣ Understanding and correctly classifying 2data types is crucial for data to behave as expected.

::: reminder
R's 6 basic data types/classes:

-   `character`: strings or characters, always quoted.
-   `numeric`: real numbers, including decimals.
-   `integer`: whole numbers.
-   `logical`: `TRUE` or `FALSE` values.
-   `factor`: categorical variables.
-   `Date/POSIXct`: dates and times.
:::

‣ Recall our dataset: 5 character variables and 9 numeric variables.

```{r}
str(non_adherence_recoded)
```

‣ Looking at our data, the only true numerical variables are `age_at_art_initation`, `bmi_initiation_art`, `cd4_initiation_art`, and `nr_of_pills_day`. Let's change all the others to factor variables using the `as.factor()` function!

‣ Change all others to factor variables using as.factor within across.

```{r}
______________ %>%
  mutate(_______)
```

‣ This should result in correct classification as expected.

::: r-practice
### Q: Changing data types {.unlisted .unnumbered}

Convert the variables in positions 13 to 29 in the `typhoid` dataset to factor.

:::


# Learning Objectives

By the end of this lesson, you will be able to:

‣ Understand how to clean column names, both automatically and manually.

‣ Eliminate duplicate entries.

‣ Correct and fix string values in your data.

‣ Convert data types as required.

# Wrap Up!

Congratulations on completing the two-part lesson on the data cleaning pipeline! You are now better equipped to tackle the cleaning of real-world datasets.

Keep practicing!

# Answer Key {.unnumbered}

### Q: Automatic cleaning {.unlisted .unnumbered}

```{r render=reactable_5_rows}
clean_names(typhoid)
```

### Q: Complete cleaning of column names {.unlisted .unnumbered}

```{r}
typhoid %>%
  clean_names() %>%
  rename_with(.fn = ~ str_replace_all(.x, pattern = "or_|of", replacement = "_")) %>%
  names()
```

### Q: Removing duplicates {.unlisted .unnumbered}

```{r render=reactable_5_rows}
# Identify duplicates
get_dupes(typhoid)

# Remove duplicates
typhoid_distinct <- typhoid %>% 
  distinct()

# Ensure all distinct rows left 
get_dupes(typhoid_distinct)
```

### Q: Transforming to lowercase {.unlisted .unnumbered}

```{r render=reactable_5_rows}
typhoid %>% 
  mutate(across(where(is.character),
                ~ tolower(.x)))
```

### Q: Fixing strings {.unlisted .unnumbered}

```{r render=reactable_5_rows}
typhoid %>%
  mutate(Householdmembers = case_match(Householdmembers, "01-May" ~ "1-5", .default=Householdmembers)) %>% 
  count(Householdmembers)
```

### Q: Changing data types {.unlisted .unnumbered}

```{r render=reactable_5_rows}
typhoid %>%
  mutate(across(13:29, ~as.factor(.)))
```

# Contributors {.unlisted .unnumbered}

The following team members contributed to this lesson:

`r tgc_contributors_list(ids = c("amckinley", "kendavidn", "lolovanco", "elmanuko"))`
