---
title: 'Working with Strings in R'
output:
  html_document:
    number_sections: true
    toc: true
    toc_float: true
    css: !expr here::here("global/style/style.css")
    highlight: kate
editor_options: 
  chunk_output_type: inline
---

```{r, echo = F, message = F, warning = F}
knitr::opts_chunk$set(fig.width = 4.5, fig.height = 3, warning = T, message = F)

# Load packages 
if(!require(pacman)) install.packages("pacman")
pacman::p_load(rlang, tidyverse, knitr, here, janitor,  reactable, gt, flextable, gtsummary, lubridate, medicaldata)

## functions
source(here::here("global/functions/misc_functions.R"))

## default render
registerS3method("reactable_5_rows", "data.frame", reactable_5_rows)
#knitr::opts_chunk$set(class.source = "tgc-code-block", render = reactable_5_rows)

```

# Introduction

Proficiency in string manipulation is a vital skill for data scientists. Tasks like cleaning messy data and formatting outputs rely heavily on the ability to parse, combine, and modify character strings. This lesson focuses on techniques for working with strings in R, utilizing functions from the {stringr} package in the tidyverse. Let's dive in!

# Learning Objectives

-   Understand the concept of strings and rules for defining them in R

-   Use escapes to include special characters like quotes within strings

-   Employ {stringr} functions to format strings:

    -   Change case with `str_to_lower()`, `str_to_upper()`, `str_to_title()`
    -   Trim whitespace with `str_trim()` and `str_squish()`
    -   Pad strings to equal width with `str_pad()`
    -   Wrap text to a certain width using `str_wrap()`

-   Split strings into parts using `str_split()` and `separate()`

-   Combine strings together with `paste()` and `paste0()`

-   Extract substrings from strings using `str_sub()`

# Packages

```{r, warning = FALSE, message = FALSE, echo = TRUE}
# Loading required packages
if(!require(pacman)) install.packages("pacman")
pacman::p_load(______, ______, ______)
```

## Defining Strings

‣ Character strings in R can be defined using single or double quotes.

‣ Matching starting and ending quotation marks are necessary.

```{r}
string_1 <- "Hello" # Using double quotes
string_2 <- 'Hello' # Using single quotes
```

‣ Cannot include double quotes inside a string that starts and ends with double quotes. The same applies to single quotes inside a string that starts and ends with single quotes.

```{r}
will_not_work <- "Double quotes " inside double quotes"
will_not_work <- 'Single quotes ' inside double quotes'
```

‣ Mixing single quotes inside double quotes, and vice versa, is allowed.

```{r}
single_inside_double <- "Single quotes ' inside double quotes"
```

‣ Use the escape character `\` to include quotes within strings.

```{r}
single_quote <- 'Single quotes \' inside sing quotes'
double_quote <- "Double quotes \" inside double quotes"
```

‣ `cat()` function is used to display strings as output.

```{r}
cat(single_quote)
cat(double_quote)
```

::: side-note
Since `\` is the escape character, you must use `\\` to include a literal backslash in a string:

```{r}
backslash <- "This is a backslash: \\"
cat(backslash)
```
:::

**PRACTICE TIME !**

::: practice
### Q: Error Spotting in String Definitions {.unlisted .unnumbered}

Below are attempts to define character strings in R, with two out of five lines containing an error. Identify and correct these errors.

```{r, eval = FALSE}
ex_a <- 'She said, "Hello!" to him.'
ex_b <- "She said \"Let's go to the moon\""
ex_c <- "They've been "best friends" for years."
ex_d <- 'Jane\\'s diary'
ex_e <- "It's a sunny day!
```
:::

## String Formatting in R with {stringr}

‣ {stringr} package helps in formatting strings for analysis and visualization.

‣ Case changes

‣ Handling whitespace

‣ Standardizing length

‣ Text wrapping

### Changing Case

‣ Standardizing strings or preparing them for display often requires case conversion.

‣ `str_to_upper()` converts strings to uppercase.

```{r}
str_to_upper("hello world") 
```

‣ `str_to_lower()` converts strings to lowercase.

```{r}
str_to_lower("Goodbye")
```

‣ `str_to_title()` capitalizes the first letter of each word. Ideal for titling names, subjects, etc.

```{r}
str_to_title("string manipulation")
```

### Handling Whitespace

‣ Strings can be made neat and uniform by managing whitespace.

‣ Use `str_trim()` to remove leading and trailing whitespace.

```{r}
str_trim(" trimmed ")
```

‣ `str_squish()` also removes whitespace at the start and end, *and* reduces multiple internal spaces to one.

```{r}
str_squish("   too   much    space   ") 

# notice the difference with str_trim
str_trim("   too   much    space   ") 
```

## Text Padding

‣ `str_pad()` is used to pad strings to a specified width.

‣ It helps to standardize the length of strings by adding characters.

```{r}
# Pad the number "7" on the left to a total width of 3 with "0"
str_pad(______, ______, ______)
```

‣ first argument: the string to pad

‣ `width` sets the final string width, `pad` specifies the padding character.

‣ `side` argument can be "left", "right", or "both".

‣ on the right:

```{r}
# Pad the number "7" on the right to length 4 with "_"
str_pad("7", width = 4, side = "right", pad = "_")
```

‣ on both sides:

```{r}
# Pad the number "7" on both sides to a total width of 5 with "_"
str_pad(______, width = __, side = "______", pad = "______") 
```

## Text Wrapping

‣ `str_wrap()` wraps text to fit a set width, useful for confined spaces.

```{r}
example_string <- "String Manipulation with str_wrap can enhance readability in plots."
wrapped_to_10 <- str_wrap(______, width = ______)
wrapped_to_10
```

‣ `cat()` displays strings with line breaks, making them readable.

```{r}
cat(______)
```

‣ Setting the width to 1 essentially splits the string into individual words:

```{r}
cat(str_wrap(wrapped_to_10, width = 1))
```

‣ Here's an example of using `str_wrap()` in ggplot2 for neat titles:

```{r}
long_title <- "This is an example of a very long title, which would usually run over the end of your ggplot, but you can wrap it with str_wrap to fit within a specified character limit."

# Example plot without title wrapping
ggplot(women, aes(height, weight)) +
  geom_point() +
  labs(title = long_title)

# Now, add wrapped title at 80 characters
ggplot(women, aes(height, weight)) +
  geom_point() +
  labs(title = str_wrap(______, width = ______))
```

**PRACTICE TIME !**

::: practice
### Q: Cleaning Patient Name Data {.unlisted .unnumbered}

A dataset contains patient names with inconsistent formatting and extra white spaces. Use the {stringr} package to standardize this information:

```{r, eval = FALSE}
patient_names <- c("  john doe", "ANNA SMITH   ", "Emily Davis")
# 1. Trim white spaces from each name.
# 2. Convert each name to title case for consistency.
```
:::

::: practice
### Q: Standardizing Drug Codes {.unlisted .unnumbered}

The following (fictional) drug codes are inconsistently formatted. Standardize them by padding with zeros to ensure all codes are 8 characters long:

```{r, eval = FALSE}
drug_codes <- c("12345", "678", "91011")
# Pad each code with zeros on the left to a fixed width of 8 characters.
```
:::

::: practice
### Q: Wrapping Medical Instructions {.unlisted .unnumbered}

Use `str_wrap()` to format the following for better readability:

```{r}
instructions <- "Take two tablets daily after meals. If symptoms persist for more than three days, consult your doctor immediately. Do not take more than the recommended dose. Keep out of reach of children."

ggplot(data.frame(x = 1, y = 1), aes(x, y, label = instructions)) +
  geom_label() + 
  theme_void()

# Now, wrap the instructions to a width of 50 characters then plot again.


```
:::

## Applying String Formatting to a Dataset

‣ We'll learn to clean and standardize data using {stringr} functions.

‣ Our focus: a dataset on **HIV care** in **Zambézia Province**, Mozambique.

‣ The dataset contains **formatting inconsistencies** intentionally added for learning.

```{r}
# Load the messy dataset
hiv_dat_messy_1 <- openxlsx::read.xlsx(here("data/hiv_dat_messy_1.xlsx")) %>%
  as_tibble()

# Observe the formatting issues in these columns
hiv_dat_messy_1 %>%
  select(______, ______, ______, ______)
```

‣ Use `tabyl` to count and identify unique values, highlighting inconsistencies.

```{r}
# Unique value counts for spotting inconsistencies
hiv_dat_messy_1 %>% tabyl(health_unit)
hiv_dat_messy_1 %>% tabyl(education)
hiv_dat_messy_1 %>% tabyl(regimen)
hiv_dat_messy_1 %>% tabyl(district)
```

‣ `tbl_summary` from {gtsummary} visualizes casing, spacing, and format issues.

```{r}
# Summarize data to view inconsistencies before cleaning
hiv_dat_messy_1 %>%
  select(district, health_unit, education,regimen) %>%
  tbl_summary()
```

‣ Next, we systematically clean each variable for consistency.

```{r}
# Apply cleaning functions to standardize data
hiv_dat_clean_1 <- hiv_dat_messy_1 %>%
  mutate(
    district = str_to_title(str_trim(______)), # Standardize district names
    health_unit = str_squish(______),          # Remove extra spaces
    education = str_to_title(______),          # Standardize education levels
    regimen = str_to_upper(______)             # Regimen column consistency
  )
```

‣ Confirm improvements by re-running `tbl_summary()`.

```{r}
# Check the cleaned data
hiv_dat_clean_1 %>%
  select(district, health_unit, education, regimen) %>%
  tbl_summary()
```

‣ Address plotting issues with `ggplot` due to lengthy `health_unit` labels.

```{r}
ggplot(hiv_dat_clean_1, aes(x = health_unit)) +
  geom_bar()
```

```{r}
# Use str_wrap to adjust label lengths for better plot display
______ %>%
  ggplot(aes(x = str_wrap(______, width = ______))) 
  geom_bar()
```

‣ Refine the plot by correcting the axis title.

```{r}
# Finalize plot adjustments
hiv_dat_clean_1 %>% 
  ggplot(aes(x = str_wrap(health_unit, width = 20))) +
  geom_bar() +
  labs(______)
```

**PRACTICE TIME!**

::: practice
### Q: Formatting a Tuberculosis Dataset {.unlisted .unnumbered}

In this exercise, you will clean a dataset, `lima_messy`, originating from a tuberculosis (TB) treatment adherence study in Lima, Peru. More details about the study and the dataset are available [here](https://zenodo.org/records/4992464){target="_blank"}.

Begin by importing the dataset:

```{r}
lima_messy_1 <- openxlsx::read.xlsx(here("data/lima_messy_1.xlsx")) %>% 
  as_tibble()
lima_messy_1
```

Your task is to clean the `marital_status`, `sex`, and `age` variables in `lima_messy`. Following the cleaning process, generate a summary table using the `tbl_summary()` function. Aim for your output to align with this structure:

| **Characteristic**     | **N = 1,293** |
|------------------------|---------------|
| marital_status         |               |
| Divorced / Separated   | 93 (7.2%)     |
| Married / Cohabitating | 486 (38%)     |
| Single                 | 677 (52%)     |
| Widowed                | 37 (2.9%)     |
| sex                    |               |
| F                      | 503 (39%)     |
| M                      | 790 (61%)     |
| age                    |               |
| 21 and younger         | 338 (26%)     |
| 22 to 26               | 345 (27%)     |
| 27 to 37               | 303 (23%)     |
| 38 and older           | 307 (24%)     |

Implement the cleaning and summarize:

```{r eval = F}
# Create a new object for cleaned data
lima_clean <- lima_messy %>%
  mutate(
    # Clean marital_status
    
    # Clean sex
    
    # Clean age
    
  )

# Check cleaning
lima_clean %>% 
  select(marital_status, sex, age) %>% 
  tbl_summary()
```
:::

::: practice
### Q: Wrapping Axis Labels in a Plot {.unlisted .unnumbered}

Using the cleaned dataset `lima_clean` from the previous task, create a bar plot to display the count of participants by `marital_status`. Then wrap the axis labels on the x-axis to a maximum of 15 characters per line for readability.

```{r eval = F}
# Create your bar plot with wrapped text here:

```
:::

## Splitting Strings

‣ Common data manipulation tasks include splitting and combining strings.

‣ `stringr::str_split()` and `tidyr::separate()` are tidyverse functions for this purpose.

### Using `str_split()`

‣ `str_split()` divides strings into parts.

‣ To split `example_string` at each hyphen:

```{r}
example_string <- "split-this-string"
str_split(______, ______)
```

‣ Direct application to a dataframe is complex.

‣ With IRS dataset, focus on `start_date_long`:

```{r}
irs <- read_csv(here("data/Illovo_data.csv"))
irs_dates_1 <- irs %>% select(village, start_date_long)
irs_dates_1
```

‣ To extract month, day, and year from `start_date_long`:

```{r}
irs_dates_1 %>% 
  mutate(start_date_parts = str_split(______, ______))
```

‣ For readability, use `unnest_wider()`:

```{r}
irs_dates_1 %>% 
  mutate(start_date_parts = str_split(start_date_long, ______)) %>% 
  unnest_wider(______, ______)
```

### Using `separate()`

‣ `separate()` is more straightforward for splitting.

‣ To split into `month`, `day`, `year`:

```{r}
irs_dates_1 %>%
  separate(start_date_long, into = c("month", "day", "year"), sep = " ")
```

‣ the `separate()` requires specifying:

-   the column to be split

-   into: names of the new columns

-   sep: separator character

‣ To keep the original column:

```{r}
irs_dates_1 %>%
  separate(start_date_long, into = c("month", "day", "year"), sep = " ", ______)
```

::: side-note
Alternatively, the `lubridate` package offers functions to extract date components:

```{r}
irs_dates_1 %>% 
  mutate(start_date_long = mdy(start_date_long)) %>% 
  mutate(day = day(start_date_long),
         month = month(start_date_long, label = TRUE),
         year = year(start_date_long))
```
:::

‣ If rows miss parts, `separate()` warns

‣ Demonstrating with dates missing "April":

```{r}
irs_dates_with_problem <- 
  irs_dates_1 %>% 
  mutate(start_date_missing = str_replace(start_date_long, "April ", ""))
irs_dates_with_problem
```

‣ Splitting with missing parts:

```{r warning=TRUE}
irs_dates_with_problem %>% 
  separate(start_date_missing, into = c("month", "day", "year"), sep = " ")
```

‣ Now we have the day and month in the wrong columns for some rows.

::: practice
### Q: Splitting Age Range Strings {.unlisted .unnumbered}

Consider the `esoph_ca` dataset, from the {medicaldata} package, which involves a case-control study of esoph ageal cancer in France.

```{r}
medicaldata::esoph_ca %>% as_tibble()
```

Split the age ranges in the `agegp` column into two separate columns: `agegp_lower` and `agegp_upper`.

After using the `separate()` function, the "75+" age group will require special handling. Use `readr::parse_number()`or another method to convert the lower age limit ("75+") to a number.

```{r eval = F}
medicaldata::esoph_ca %>%
  separate(________) %>% 
  # convert 75+ to a number 
  mutate(________)
```
:::

## Separating Special Characters

‣ To use the `separate()` function on special characters (., +, \*, ?) need to be escaped in `\\`

\
‣ Consider the scenario where dates are formatted with periods:

```{r}
# Correct separation of dates with periods
irs_with_period <- irs_dates_1 %>% 
  mutate(start_date_long = format(lubridate::mdy(start_date_long), "%d.%m.%Y"))
irs_with_period
```

‣ When attempting to separate this date format directly with `sep = "."` :

```{r}
irs_with_period %>%
  separate(start_date_long, into = c("day", "month", "year"), sep = ______ )
```

‣ This is because, in regex (regular expressions), the period is a special character.

‣ The correct approach is to escape the period uses a double backslash (\\):

```{r}
irs_with_period %>% 
  separate(start_date_long, into = c("day", "month", "year"), sep = ______)
```

‣ Now, the function understands to split the string at each literal period.

‣ When using other special characters like `+`, `*`, or `?`, they need to be preceded with a double backslash (\\) in the `sep` argument.

::: side-note
**What is a Special Character?**

In regular expressions, which help find patterns in text, special characters have specific roles. For example, a period (.) is a wildcard that can represent any character. So, in a search, "do.t" could match "dolt," "dost," or "doct" Similarly, the plus sign (+) is used to indicate one or more occurrences of the preceding character. For example, "ho+se" would match "hose" or "hooose" but not "hse." When we need to use these characters in their ordinary roles, we use a double backslash (`\\`) before them, like "`\\.`" or "`\\+.`" More on these special characters will be covered in a future lesson.
:::

::: practice
### Q: Separating Special Characters {.unlisted .unnumbered}

Your next task involves the `hiv_dat_clean_1` dataset. Focus on the `regimen` column, which lists drug regimens separated by a `+` sign. Your goal is to split this column into three new columns: `drug_1`, `drug_2`, and `drug_3` using the `separate()` function. Pay close attention to how you handle the `+` separator. Here's the column:

```{r}
hiv_dat_clean_1 %>% 
  select(regimen)
```
:::

# Combining Strings with `paste()`

‣ Concatenate strings with `paste()`

‣ To combine two simple strings:

```{r}
string1 <- "Hello"
string2 <- "World"
paste(string1, string2)
```

‣ Let's demonstrate this with the IRS data.

‣ First, we'll separate the start date into individual columns:

```{r}
irs_dates_separated <- # store for later use
  irs_dates_1 %>%
  separate(start_date_long, into = c("month", "day", "year"), sep = " ", remove = FALSE)
irs_dates_separated
```

‣ Then, recombine day, month and year with `paste()`:

```{r}
irs_dates_separated %>%
  select(day, month, year) %>%
  mutate(start_date_long_2 = paste(______, ______, ______))
```

‣ `sep` argument specifies the separator between elements

‣ For different separators, we can write:

```{r}
irs_dates_separated %>%
  mutate(start_date_long_2 = paste(______, ______, ______, ______))
```

‣ To concatenate without spaces, we can set `sep = ""`:

```{r}
irs_dates_separated %>%
  select(day, month, year) %>%
  mutate(start_date_long_2 = paste(day, month, year, sep = ______))
```

‣ Or use `paste0()` function, which is equivalent to `paste(..., sep = "")`:

```{r}
irs_dates_separated %>%
  select(day, month, year) %>%
  mutate(start_date_long_2 = paste0(day, month, year))
```

‣ Combine `paste()` with other string functions to solve a realistic data problem.

‣ Consider the ID column in the `hiv_dat_messy_1` dataset:

```{r}
hiv_dat_messy_1 %>% 
  select(patient_id)
```

‣ Standardize these IDs to the same number of characters.

‣ Use `separate()` to split the IDs into parts, then use `paste()` to recombine them:

```{r}
hiv_dat_messy_1 %>%
  select(patient_id) %>% # for visibility
  separate(patient_id, into = c(______, ______), sep = ______, remove = ______) %>%
  mutate(patient_num = str_pad(patient_num, width = 5, side = "left", pad = "0")) %>%
  mutate(patient_id_padded = paste(______, ______, sep = ______))
```

‣ In this example, `patient_id` is split into a prefix and a number.\
‣ The number is padded with zeros to ensure consistent length\
‣ They're concatenated back together using `paste()` with a hyphen as the separator.

::: practice
### Q: Standardizing IDs in the `lima_messy_1` Dataset {.unlisted .unnumbered}

In the `lima_messy_1` dataset, the IDs are not zero-padded, making them hard to sort.

For example, the ID `pe-998` is at the top of the list after sorting in descending order, which is not what we want.

```{r}
lima_messy_1 %>%
  select(id) %>%
  arrange(desc(id)) # sort in descending order (highest IDs should be at the top)
```

Try to fix this using a similar procedure to the one used for `hiv_dat_messy_1`.

**Your Task:**

-   Separate the ID into parts.
-   Pad the numeric part for standardization.
-   Recombine the parts using `paste()`.
-   Resort the IDs in descending order. The highest ID should end in `2347`

```{r eval = F}
lima_messy_1 %>%
  ______________
```
:::

::: practice
### Q: Creating Summary Statements {.unlisted .unnumbered}

Create a column containing summary statements combining `village`, `start_date_default`, and `coverage_p` from the `irs` dataset. The statement should describe the spray coverage for each village.

**Desired Output:** "For village X, the spray coverage was Y% on Z date."

**Your Task:** - Select the necessary columns from the `irs` dataset. - Use `paste()` to create the summary statement.

```{r eval = F}
irs %>% 
  select(village, start_date_default, coverage_p) %>% 
  ______________________
```
:::

# Subsetting Strings with `str_sub`

‣ `str_sub` is used to extract parts of a string based on character positions

‣ Basic syntax: `str_sub(string, start, end)`

‣ Example: Extracting first 2 characters from patient IDs

```{r}
patient_ids <- c("ID12345-abc", "ID67890-def")
str_sub(patient_ids, __, __) 
```

‣ To extract other characters, like the first 5, adjust the `start` and `end` values

```{r}
str_sub(patient_ids, __, __)
```

‣ Negative values count backward from the string end, useful for suffixes

‣ Examples: Get the last 4 characters of patient IDs:

```{r}
str_sub(patient_ids, __, __)
```

‣ `str_sub` will not error out if indices exceed string length

```{r}
str_sub(patient_ids, __, __) # Safely returns the full string if range exceeds string length
```

‣ Within `mutate()`, `str_sub` can be used to transform columns in a data frame

‣ Example: Extracting year and month from `start_date_default` column and create a new column called `year_month`:

```{r}
irs %>% 
  select(______) %>%
  mutate(______ = str_sub(______, start = __, end = __))
```

**PRACTICE TIME!**

::: practice
### Q: Extracting ID Substrings {.unlisted .unnumbered}

Use `str_sub()` to isolate just the numeric part of the `patient_id` column in the `hiv_dat_messy_1` dataset.

```{r eval = F}
hiv_dat_messy_1 %>% 
  select(patient_id) %>% 
  # Complete the code below:
  mutate(numeric_id = str_sub(patient_id, __, __))
```
:::

# Wrap Up! {.unnumbered}

Congratulations on reaching the end of this lesson! You've learned about strings in R and various functions to manipulate them effectively.

The table below gives a quick recap of the key functions we covered. Remember, you don't need to memorize all these functions. Knowing they exist and how to look them up (like using Google) is more than enough for practical applications.

+------------------+---------------------------------------------------------------+------------------------------------------+---------------------+
| Function         | Description                                                   | Example                                  | Example Output      |
+------------------+---------------------------------------------------------------+------------------------------------------+---------------------+
| `str_to_upper()` | Convert characters to uppercase                               | `str_to_upper("hiv")`                    | "HIV"               |
+------------------+---------------------------------------------------------------+------------------------------------------+---------------------+
| `str_to_lower()` | Convert characters to lowercase                               | `str_to_lower("HIV")`                    | "hiv"               |
+------------------+---------------------------------------------------------------+------------------------------------------+---------------------+
| `str_to_title()` | Convert first character of each word to uppercase             | `str_to_title("hiv awareness")`          | "Hiv Awareness"     |
+------------------+---------------------------------------------------------------+------------------------------------------+---------------------+
| `str_trim()`     | Remove whitespace from start & end                            | `str_trim(" hiv ")`                      | "hiv"               |
+------------------+---------------------------------------------------------------+------------------------------------------+---------------------+
| `str_squish()`   | Remove whitespace from start & end and reduce internal spaces | `str_squish(" hiv cases ")`              | "hiv cases"         |
+------------------+---------------------------------------------------------------+------------------------------------------+---------------------+
| `str_pad()`      | Pad a string to a fixed width                                 | `str_pad("45", width = 5)`               | "00045"             |
+------------------+---------------------------------------------------------------+------------------------------------------+---------------------+
| `str_wrap()`     | Wrap a string to a given width (for formatting output)        | `str_wrap("HIV awareness", width = 5)`   | "HIV\nawareness"    |
+------------------+---------------------------------------------------------------+------------------------------------------+---------------------+
| `str_split()`    | Split elements of a character vector                          | `str_split("Hello-World", "-")`          | c("Hello", "World") |
+------------------+---------------------------------------------------------------+------------------------------------------+---------------------+
| `paste()`        | Concatenate vectors after converting to character             | `paste("Hello", "World")`                | "Hello World"       |
+------------------+---------------------------------------------------------------+------------------------------------------+---------------------+
| `str_sub()`      | Extract and replace substrings from a character vector        | `str_sub("HelloWorld", 1, 4)`            | "Hell"              |
+------------------+---------------------------------------------------------------+------------------------------------------+---------------------+
| `separate()`     | Separate a character column into multiple columns             | `separate(tibble(a = "Hello-World"), a,` | `|b     |c     |`\  |
|                  |                                                               |                                          | `|Hello |World |`   |
|                  |                                                               | `into = c("b", "c"),`                    |                     |
|                  |                                                               |                                          |                     |
|                  |                                                               | `sep = "-")`                             |                     |
+------------------+---------------------------------------------------------------+------------------------------------------+---------------------+

Note that while these functions cover common tasks such as string standardization, splitting and joining strings, this introduction only scratches the surface of what's possible with the {stringr} package. If you work with a lot of raw text data, you may want to do further exploring on the [stringr](https://stringr.tidyverse.org){target="_blank"} website.

# Answer Key {.unnumbered}

### Q: Error Spotting in String Definitions {.unlisted .unnumbered}

1.  **`ex_a`**: Correct.
2.  **`ex_b`**: Correct.
3.  **`ex_c`**: Error. Corrected version: `ex_c <- "They've been \"best friends\" for years."`
4.  **`ex_d`**: Error. Corrected version: `ex_d <- 'Jane\'s diary'`
5.  **`ex_e`**: Error. Close quote missing. Corrected version: `ex_e <- "It's a sunny day!"`

### Q: Cleaning Patient Name Data {.unlisted .unnumbered}

```{r eval = F}
patient_names <- c("  john doe", "ANNA SMITH   ", "Emily Davis")

patient_names <- str_trim(patient_names) # Trim white spaces
patient_names <- str_to_title(patient_names) # Convert to title case
```

### Q: Standardizing Drug Codes {.unlisted .unnumbered}

```{r eval = F}
drug_codes <- c("12345", "678", "91011")

# Pad each code with zeros on the left to a fixed width of 8 characters.
drug_codes_padded <- str_pad(drug_codes, 8, pad = "0")
```

### Q: Wrapping Medical Instructions {.unlisted .unnumbered}

```{r eval = F}
instructions <- "Take two tablets daily after meals. If symptoms persist for more than three days, consult your doctor immediately. Do not take more than the recommended dose. Keep out of reach of children."

# Wrap instructions
wrapped_instructions <- str_wrap(instructions, width = 50)

ggplot(data.frame(x = 1, y = 1), aes(x, y, label = wrapped_instructions)) +
  geom_label() + 
  theme_void()
```

### Q: Formatting a Tuberculosis Dataset {.unlisted .unnumbered}

The steps to clean the `lima_messy` dataset would involve:

```{r eval = F}
lima_clean <- lima_messy %>%
  mutate(
    marital_status = str_squish(str_to_title(marital_status)), # Clean and standardize marital_status
    sex = str_squish(str_to_upper(sex)),                        # Clean and standardize sex
    age = str_squish(str_to_lower(age))                         # Clean and standardize age
  )


lima_clean %>% 
  select(marital_status, sex, age) %>%
  tbl_summary()
```

Then, use the `tbl_summary()` function to create the summary table.

### Q: Wrapping Axis Labels in a Plot {.unlisted .unnumbered}

```{r eval = F}
# Assuming lima_clean is already created and contains marital_status
ggplot(lima_clean, aes(x = str_wrap(marital_status, width = 15))) +
  geom_bar() + 
  labs(x = "Marital Status")
```

### Q: Splitting Age Range Strings {.unlisted .unnumbered}

```{r eval = F}
esoph_ca %>% 
  select(agegp) %>% # for illustration
  separate(agegp, into = c("agegp_lower", "agegp_upper"), sep = "-") %>% 
  mutate(agegp_lower = readr::parse_number(agegp_lower))
```

### Q: Creating Summary Statements {.unlisted .unnumbered}

```{r eval = F}
irs %>% 
  select(village, start_date_default, coverage_p) %>% 
  mutate(summary_statement = paste0("For village ", village, ", the spray coverage was ", coverage_p, "% on ", start_date_default))
```

### Q: Extracting ID Substrings {.unlisted .unnumbered}

```{r eval = F}
hiv_dat_messy_1 %>% 
  select(patient_id) %>% 
  mutate(numeric_part = str_sub(patient_id, 4))
```

# Contributors {.unlisted .unnumbered}

The following team members contributed to this lesson:

`r tgc_contributors_list(ids = c("camille", "kendavidn"))`
